---
title: "Interoperability between SpatialExperiment and AnnData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interoperability between SpatialExperiment and AnnData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(renovae)

Sys.setenv(TQDM_DISABLE = "1")
```

This vignette demonstrates the seamless interoperability between the R `SpatialExperiment`
class and the Python `AnnData` format, specifically in the context of the
[novae](https://www.nature.com/articles/s41592-025-02899-6) foundation model.

## Background

The `renovae` package provides an interface to the `novae` deep learning framework for spatial transcriptomics. While
`novae` is natively implemented in Python and expects `AnnData` objects as input, `renovae` leverages `anndataR`
to allow R users to work directly with `SpatialExperiment` objects.

## The Dataset

We use a crop of the Janesick et al. (2023) Xenium human breast cancer dataset.

```{r load-data}
# Load as SpatialExperiment
spe <- load_janesick_spe()
spe

# Load as AnnData
ad <- load_janesick_ad()
ad
```

## Running the Inference

The `novae_run` function is designed to handle both input types transparently.

### Starting from AnnData

If you already have an `AnnData` object (e.g., loaded via `reticulate` or `anndataR`), you can run the model
directly. The embeddings will be returned as a matrix:

```r
out_ad <- novae_run(ad)
```

### Starting from SpatialExperiment

For R users, the most common entry point is a `SpatialExperiment`. `renovae` automatically handles the conversion to
`AnnData` behind the scenes.

```{r run-spe}
out_spe <- novae_run(spe)

# The result contains the 64-dimensional cell embeddings computed by the model
dim(out_spe)
```
