---
title: "Reproduction of SpatialExperiment-to-AnnData conversion issue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproduction of SpatialExperiment-to-AnnData conversion issue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(renovae)
```

This vignette documents an issue encountered when converting `SpatialExperiment` objects back to `AnnData`
using the `anndataR` package, specifically in the context of spatial transcriptomics data and the
[novae](https://www.nature.com/articles/s41592-025-02899-6) foundation model.

## Background

The `renovae` package provides an interface to the `novae` deep learning framework for spatial transcriptomics. While
`novae` is natively implemented in Python and expects `AnnData` objects as input, `renovae` aims to allow R users
to work directly with the native `SpatialExperiment` class.

## The Dataset

We start from an `AnnData` object representing a 25,000 x 25,000 pixel crop of the original Janesick et al. (2023)
Xenium human breast cancer dataset. In the context of the package, the original object can be loaded with:

```{r load-ad, eval=FALSE}
# This requires a Python environment with scanpy and novae installed
ad <- load_janesick_ad()
```

We used a fork of the `anndataR` package (from `drighelli/anndataR@from_as_spe`) to convert this object into the
`SpatialExperiment` native R class. This object is also provided by the package:

```{r load-spe}
spe <- load_janesick_spe()
spe
```

## Running the Inference

In principle, we could now use either to run the inference from the `novae` model. For instance, using the
`novae_run` function provided by this package, one could run the model directly on the `AnnData` object:

```r
out_ad <- novae_run(ad)
```

This returns the cell embeddings computed by the model for later inspection.

While `novae` requires `AnnData` as an input, we could start from an `SpatialExperiment` by using the conversion
function provided by [anndataR](https://github.com/scverse/anndataR) (`anndataR::from_SpatialExperiment`). We
implemented this approach in the `novae_run` helper to facilitate seamless integration for R users.

```{r, error = TRUE}
out_spe <- novae_run(spe)
```

## Observed Issue

When running `novae_run` starting from the `SpatialExperiment` object, the initial phase of the `novae` analysis
(neighbor graph computation) appears to run, but the subsequent step (representation computation) complains about
missing cell neighbor information.

We believe this issue relates to the `SpatialExperiment` -> `AnnData` conversion. Specifically, the round-trip
conversion may not be preserving the spatial neighbor structures or the underlying spatial coordinate metadata in a
way that is fully compatible with the downstream Python tools. This indicates that while the conversion provides
an `AnnData` object, it might be lacking some critical elements required by the `novae` framework's internal
processing logic.
